---
output:
  github_document:
    html_preview: false
---

<!-- README.md is generated from README.Rmd. Please edit README.Rmd. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# RGlimpse2

[![R-CMD-check](https://github.com/RGenomicsETL/RGlimpse2/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/RGenomicsETL/RGlimpse2/actions/workflows/R-CMD-check.yaml)
[![pkgdown](https://github.com/RGenomicsETL/RGlimpse2/actions/workflows/pkgdown.yaml/badge.svg)](https://rgenomicsetl.github.io/RGlimpse2/)
[![R-universe](https://RGenomicsETL.r-universe.dev/RGlimpse2/badges/version)](https://RGenomicsETL.r-universe.dev/RGlimpse2)

`RGlimpse2` runs GLIMPSE2 genotype imputation and phasing for low-coverage
whole-genome sequencing from R. It builds and installs the pinned
`GLIMPSE2_chunk`, `GLIMPSE2_split_reference`,
`GLIMPSE2_phase`, and `GLIMPSE2_ligate` programs, then invokes one child
executable per operation through direct `processx` argument vectors.

## Installation

Install development builds from the RGenomicsETL R-universe:

```{r installation}
install.packages(
  "RGlimpse2",
  repos = c(
    RGenomicsETL = "https://rgenomicsetl.r-universe.dev",
    CRAN = "https://cloud.r-project.org"
  )
)
```

Source installation requires GNU make, a C++17 compiler, and the Boost
`iostreams`, `program_options`, and `serialization` libraries. Rtools45 ships
these requirements for Windows. Linux distributions and Homebrew provide them
as ordinary development packages.

## Packaged genetic maps

GRCh37 and GRCh38 autosomal, X non-PAR, X PAR1, and X PAR2 maps are installed
with the package. Minimal zero-recombination coordinate maps are also provided
for Y non-PAR and mitochondrial sequence:

```{r maps}
library(RGlimpse2)

rglimpse2_genetic_map("GRCh38", "22")
rglimpse2_genetic_map("GRCh38", "X", region = "par1")
rglimpse2_genetic_map("GRCh37", "Y")
rglimpse2_genetic_map("GRCh38", "Y")
rglimpse2_genetic_map("GRCh37", "chrM")
rglimpse2_genetic_map("GRCh38", "MT")
head(rglimpse2_genetic_maps())
```

| Assembly | Map | Coordinates |
|---|---|---:|
| GRCh37 | Y non-PAR | 2,649,521-59,034,049 |
| GRCh38 | Y non-PAR | 2,781,480-56,887,902 |
| GRCh37 | mitochondrial (`M`, `MT`, `chrM`, or `chrMT`) | 1-16,569 |
| GRCh38 | mitochondrial (`M`, `MT`, `chrM`, or `chrMT`) | 1-16,569 |

The Y files are explicitly named `chrY_nonpar.*.gmap.gz`. The Y and
mitochondrial files contain two anchors at `0` cM because GLIMPSE requires at
least two map entries. The native conformance tests run small haploid BCF
reference/target pairs through chunk, split-reference, phase, and ligate using
`Y`, `chrY`, `M`, `MT`, `chrM`, and `chrMT`. This proves contig-label mechanics,
not scientific calibration of Y or mitochondrial imputation.

## One explicit operation at a time

Resolve the installed executables and run an operation with explicit inputs,
outputs, interval, seed, and resources:

```{r interface}
library(RGlimpse2)

executables <- rglimpse2_executables(phase_backend = "auto")
if (rglimpse2_is_error(executables)) {
  stop(executables@message)
}

genetic_map <- rglimpse2_genetic_map("GRCh38", "22")

chunks <- rglimpse2_chunk(
  input_sites = "/data/reference.sites.bcf",
  region = "chr22",
  output_chunks = "/work/chunks.chr22.txt",
  executable = executables@chunk,
  genetic_map = genetic_map,
  seed = 20260727L,
  threads = 2L
)
```

The split-reference, phase, and ligate wrappers follow the same contract. They
do not search `PATH`, overwrite an output, build shell command strings, or
retain workflow state. Invalid calls signal typed
`rglimpse2_contract_violation` conditions; expected operational failures are
returned as typed S7 error values.

## One htslib authority

The package does not vendor or discover another htslib. During configuration it
requires:

```{r htslib-contract}
Rduckhts::rduckhts_htslib_config(validate = TRUE)
```

Every executable is built with the returned headers and exact library contract.
Runtime executable discovery revalidates the packaged build version against the
htslib version supplied and loaded by `Rduckhts`.

## Runtime phase dispatch

`GLIMPSE2_phase` is built as complete, process-isolated executables:

- x86_64: scalar SIMDe, AVX2, and AVX-512F/BW/VL;
- ARM: scalar SIMDe and NEON.

A baseline C probe checks CPU features and operating-system vector state before
`"auto"` selects a binary. Explicit backend selection supports conformance and
reproducibility testing without process-global dispatch state.

```{r simd}
rglimpse2_simd_info()
scalar <- rglimpse2_executables(phase_backend = "scalar")
```

The native tests run all four packaged GLIMPSE2 programs against eight tiny
`vcfppR`-generated and `Rduckhts`-indexed BCF reference/target pairs. They cover
`1`/`chr1`, `Y`/`chrY`, and `M`/`MT`/`chrM`/`chrMT`, including haploid Y and
mitochondrial targets, and require identical fixed-seed phase records from every
runtime-supported backend. No executable test doubles are used.

## Building this fork

The upstream GLIMPSE2 tree remains the repository root. The R package is nested
under `RGlimpse2/`:

```sh
make r-package-bootstrap
make r-package-check
```

Reference-panel copying evidence is an inference diagnostic. `RGlimpse2` does
not call it local ancestry; that claim requires a separately calibrated and
validated decoder.
